Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-22 Paper/manuscript_v2/manuscript_07_01.tex
added some acks

The remainder of the paper is structured as follows. In Section~\ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section~\ref{ch5_sec_method}, the reduced-rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section~\ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit exact and approximate GPs to real and simulated data are provided in Section~\ref{ch5_sec_cases}. A brief conclusion of the work is made in Section~\ref{ch5_sec_conclusion}. Appendix~\ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. Online supplemental material with more case studies illustrating the performance and applicability of the method can be found online at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

A GP is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, that is, $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_N) \right\rbrace$, is a multivariate Gaussian distribution.

@@

We assume a GP prior for $f \sim \GP(\mu(\bm{x}), k(\bm{x}, \bm{x}'))$, where $\mu: {\rm I\!R}^D \to {\rm I\!R}$ and $k: {\rm I\!R}^D \times {\rm I\!R}^D \to {\rm I\!R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\bm{x}) &= \mathbb{E}\!\left[f(\bm{x})\right],\\ 
 	k(\bm{x}, \bm{x}') &= \mathbb{E}\!\left[\left( f(\bm{x}) - \mu(\bm{x}) \right)\left( f(\bm{x}') - \mu(\bm{x}') \right)\right].
\end{align*} 
%
The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution $\bm{f} \sim \Normal(\bm{\mu}, \bm{K})$, where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. In the following, we focus on zero-mean Gaussian processes, that is set $\mu(\bm{x}) = 0$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation. The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\Normal \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*}
%
where $\bm{k}_{\bm{f},\tilde{f}}$ is the covariance between $\bm{f}$ and $\tilde{f}$, and $k_{\tilde{f},\tilde{f}}$ is the prior variance of $\tilde{f}$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\bm{f}$ analytically,
% \begin{align*}
	% p(f^*| \bm{f}) = \Normal(f^*| \bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{k}_{\bm{f},f^*}  )
% \end{align*}

@@

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern($\nu$=3/2) covariance function. We considered an approximation to be a close enough when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$:
%
\begin{equation*}
 \frac{\varepsilon}{\int k(\tau) \,\mathrm{d}\tau} < 0.01.
\end{equation*}
%
Alternatively, these figures could be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be closely approximated given $m$ and $c$. We obtain the following main conclusions:

@@

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process using the Mat{\'e}rn($\nu$=3/2) covariance function with marginal variance $\alpha=1$ and lengthscale $\ell=0.15$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(\bm{0}, \sigma^2 \bm{I}) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function at the input values $x_i$, and $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, with $\bm{I}$ representing the identity matrix. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with the Mat{\'e}rn($\nu$=3/2) covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,2,\dots,n$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using MCMC sampling methods, would be its marginalized form,
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata1}
\bm{y} \sim \Normal(\bm{0},\bm{K}+ \sigma^2 \bm{I} ),
\end{equation*}
%
where the function values $\bm{f}$ have been integrated out, yielding a lower-dimensional parameter space over which to do inference, reducing the time of computation and improving the sampling and the effective number of samples.

@@

Figure~\ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the exact GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generating function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, exact GP, HSGP and spline, are similar in the interpolation regions of the input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it is seen that the HSGP method yield a good approximation of the exact GP model for both interpolation and extrapolation. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

Figure~\ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the exact GP and 10 times faster than the spline model for this particular model and data. Also, it is seen that the computation time increases slowly as a function of the number of basis functions.

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{ https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_1D-Simulated-data}}\,.
%

@@

In the online supplemental material, additional case studies are presented. From those examples, it can be seen how the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Even though, for a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the exact GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). For moderate sized datasets, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure~C.3 in the online material). However, the HSGP method will be computationally faster than the exact GP for larger datasets due the cubic scaling of the exact GP. In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.

@@

\subsection{Birthday data}\label{ch5_sec_birthday}
%
This example is an analysis of patterns in birthday frequencies in a dataset containing records of all births in the United States on each day during the period 1969–1988. The model decomposes the number of births along all the period in longer-term trend effects, patterns during the year, day-of-week effects, and special days effects. The special days effects cover patterns such as possible fewer births on Halloween, Christmas or new year, and excess of births on Valentine’s Day or the days after Christmas (due, presumably, to choices involved in scheduled deliveries, along with decisions of whether to induce a birth for health reasons). \citet{gelman2013bayesian} presented an analysis using exact GP and maximum a posteriori inference. As the total number of days within the period is $T=7305$ ($t=1,2,\dots,T$), a full Bayesian inference with MCMC for a exact GP model is memory and time consuming. We will use the HSGP method as well as the low-rank GP model with a periodic covariance function described in Appendix~\ref{ch5_sec_periodic} which is based on expanding the periodic covariance function into a series of stochastic resonators \citep{solin2014explicit}.

Let $y_t$ denote the number of births on the $t$'th day. The observation model is a normal distribution with mean function $\mu(t)$ and noise variance $\sigma^2$,
%
\begin{equation*}
y_{t} \sim \Normal(\mu(t),\sigma^2).
\end{equation*}
%
The mean function $\mu(t)$ will be defined as an additive model in the form
%
\begin{equation} \label{ch5_eq_mean_brithday}
\mu(t) = f_1(t) + f_2(t) + f_3(t) + f_4(t).
\end{equation}

The component $f_1(t)$ represents the long-term trends modeled by a GP with squared exponential covariance function,
%
\begin{equation*}
f_1(t) \sim \GP(0,k_1), \hspace{5mm} k_1(t,t') = \alpha_1 \exp\!\!\left(\!-\frac{1}{2} \frac{(t-t')^2}{\ell_1^2}\right), 
\end{equation*}
%
which means the function values $\bm{f}_1=\{f_1(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $\bm{K}_1$, where $K_{1_{t,s}}=k_1(t,s)$, with $t,s=1,\dots,T$. $\alpha_1$ and $\ell_1$ represent the marginal variance and lengthscale, respectively, of this GP prior component. 
The component $f_2(t)$ represents the yearly smooth seasonal pattern, using a periodic squared exponential covariance function (with period 365.25 to match the average length of the year) in a GP model,
%
\begin{equation*}
f_2(t) \sim \GP(0,k_2), \hspace{5mm} k_2(t,t') = \alpha_2 \exp\!\!\left(\!-\frac{2\,\sin^{\!2}\!(\pi(t-t')/365.25}{\ell_2^2}\right).
\end{equation*}
% which means the function values $\bm{f}_2=\{f_2(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_2$, where $K_{2_{t,s}}=k_2(t,s)$, with $t,s=1,\dots,T$.

The component $f_3(t)$ represents the weekly smooth pattern using a periodic squared exponential covariance function (with period 7 of length of the week) in a GP model,
%
\begin{equation*}
f_3(t) \sim \GP(0,k_3), \hspace{5mm} k_3(t,t') = \alpha_3 \exp\!\!\left(\!-\frac{2\,\sin^{\!2}\!(\pi(t-t')/7}{\ell_3^2}\right). 
\end{equation*}
% which means the function values $\bm{f}_3=\{f_3(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_3$, where $K_{3_{t,s}}=k_3(t,s)$, with $t,s=1,\dots,T$.

The component $f_4(t)$ represents the special days effects, modeled as a horse-shoe prior model \citep{carvalho2010,piironen2017sparsity}:
%
\begin{equation*}
f_4(t) \sim \Normal(0,\lambda^2_t \tau^2), \hspace{5mm} \lambda^2_t \sim \mathcal{C}^{+}(0,1).
\end{equation*}
%
A horse-shoe prior allows for sparse distributed effects. Its global parameter $\tau$ pulls all the weights (effects) globally towards zero, while the thick half-Cauchy tails for the local scales $\lambda_t$ allow some of the weights to escape the shrinkage. Different levels of sparsity can be accommodated by changing the value of $\tau$: for large $\tau$ all the variables have very diffuse prior distributions with very little shrinkage towards zero, but letting $\tau \rightarrow 0$ will shrink all the weights $f_4(t)$ to zero \citep{piironen2016hyperprior}. 
%More details about the specified priors for $\tau$ and $\lambda_t$ have been included in Appendix \ref{app_horse_shoe}.

@@

The year effects $f_2(t)$ and week effects $f_3(t)$ use a periodic covariance function  and thus do no fit under the main framework of the HSGP approximation covered in this paper. However, they do have a representation based on expanding periodic covariance functions into a series of stochastic resonators (Appendix~\ref{ch5_sec_periodic}). Thus, the functions $f_2(t)$ and $f_3(t)$ are approximated as in equation (\ref{ch5_eq_f_period}), with variance coefficients $\tilde{q}_j^2$ as in equation (\ref{ch5_eq_q_2}).
We use $J=10$ cosine terms. The lengthscale estimates $\hat{\ell}_2$ and $\hat{\ell}_3$ for the GP components $f_2(t)$ and $f_3(t)$, respectively, are bigger than the minimum lengthscale reported by Figure \ref{ch5_fig28_m_lscale_periodic} as function of the number of cosine terms $J$, which means that the approximations are good.

@@

The Stan model code for the approximate GP model of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Birthday-data}}\,.

@@

As the WBC predictor values were strictly positive and highly skewed, a logarithm transformation is used. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i \mid f_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

>>>>

The remainder of the paper is structured as follows. In Section~\ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section~\ref{ch5_sec_method}, the reduced-rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section~\ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit exact and approximate GPs to real and simulated data are provided in Section~\ref{ch5_sec_cases}. A brief conclusion of the work is made in Section~\ref{ch5_sec_conclusion}. Appendix~\ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

A GP is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, that is, $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. GPs have the defining property that the marginal distribution of any finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.

@@

We assume a GP prior for $f \sim \GP(\mu(\bm{x}), k(\bm{x}, \bm{x}'))$, where $\mu: {\rm I\!R}^D \to {\rm I\!R}$ and $k: {\rm I\!R}^D \times {\rm I\!R}^D \to {\rm I\!R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\bm{x}) &= \mathbb{E}\!\left[f(\bm{x})\right],\\ 
 	k(\bm{x}, \bm{x}') &= \mathbb{E}\!\left[\left( f(\bm{x}) - \mu(\bm{x}) \right)\left( f(\bm{x}') - \mu(\bm{x}') \right)\right].
\end{align*}

The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution $\bm{f} \sim \Normal(\bm{\mu}, \bm{K})$, where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. In the following, we focus on zero-mean Gaussian processes, that is set $\mu(\bm{x}) = 0$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation. The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\Normal \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*}
%
where $\bm{k}_{\bm{f},\tilde{f}}$ is the covariance between $\bm{f}$ and $\tilde{f}$, and $k_{\tilde{f},\tilde{f}}$ is the prior variance of $\tilde{f}$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\bm{f}$ analytically,
% \begin{align*}
	% p(f^*| \bm{f}) = \Normal(f^*| \bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{k}_{\bm{f},f^*}  )
% \end{align*}

@@

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact with each other in relation to a close approximation of the HSGP model, in the cases of a GP with square exponential covariance function and Mat\'ern($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a accurate approximation in the sense of satisfying equation (\ref{eq_diff_covs}). Similarly for Figure \ref{fig6_lscale_vs_J_vs_c_Matern} in the case of a Mat\'ern($\nu$=3/2) covariance function. We considered an approximation to be a close enough when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in equation (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$:
%
\begin{equation*}
 \frac{\varepsilon}{\int k(\tau) \,\mathrm{d}\tau} < 0.01.
\end{equation*}
%
Alternatively, these figures could be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be closely approximated given $m$ and $c$. We obtain the following main conclusions:

@@

This example consists of a simulated dataset with $n=250$ ($i=1,2,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(0, \sigma^2  I) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $I$ represents the identity matrix and $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function values to the noisy data. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with a Mat{\'e}rn($\nu$=3/2) covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using MCMC sampling methods, would be its marginalized form,
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata1}
\bm{y} \sim \Normal(0,\bm{K}+ \sigma^2 \bm{I} ),
\end{equation*}
%
where the function values $\bm{f}$ have been integrated out, yielding a lower-dimensional parameter space over which to do inference, reducing the time of computation and improving the sampling and the effective number of samples.

@@

Figure~\ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the exact GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generative function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data, which have not been taking part on training the models, are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, exact GP, HSGP and spline, are pretty similar within the interpolation input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the exact GP model for interpolating and extrapolating data. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

Figure~\ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the exact GP and 10 times faster than the spline model, for this particular application with a univariate input space. Also, the increase in computation time as a function of the number of basis functions in a univariate input space is relatively slight.

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{ https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_1D-Simulated-data}}.
%

@@

In the online supplemental material, some other case studies are presented. From those examples, it can be seen how computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Even though, in a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the exact GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). However, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure~C.3 in the online material). In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.

@@

\subsection{Birthday data}\label{ch5_sec_birthday}
%
This example is an analysis of patterns in birthday frequencies in a dataset containing records of all births in the United States on each day during the period 1969–1988. The model decomposes the number of births along all the period in longer-term trend effects, patterns during the year, day-of-week effects, and special days effects. The special days effects cover patterns such as possible fewer births on Halloween, Christmas or new year, and excess of births on Valentine’s Day or the days after Christmas (due, presumably, to choices involved in scheduled deliveries, along with decisions of whether to induce a birth for health reasons). \citet{gelman2013bayesian} presented an analysis using exact GP and maximum a posteriori inference. As the total number of days within the period is $T=7305$ ($t=1,\dots,T$), a full Bayesian inference with MCMC for a exact GP model is memory and time consuming. We will use the HSGP method as well as the low-rank GP model with a periodic covariance function described in Appendix~\ref{ch5_sec_periodic} which is based on expanding the periodic covariance function into a series of stochastic resonators \citep{solin2014explicit}.

Let us denote $y_t$ as the number of births of day $t$. The observational model is a normal model with parameters the mean function $\mu(t)$ and noise variance $\sigma^2$,
%
\begin{equation*}
y_{t} \sim \Normal(\mu(t),\sigma^2).
\end{equation*}
%
The mean function $\mu(t)$ will be defined as an additive model in the form: 
%
\begin{equation} \label{ch5_eq_mean_brithday}
\mu(t) = f_1(t) + f_2(t) + f_3(t) + f_4(t).
\end{equation}

The component $f_1(t)$ represents the long-term trends modeled by a GP with squared exponential covariance function,
%
\begin{equation*}
f_1(t) \sim \GP(0,k_1), \hspace{5mm} k_1(t,t') = \sigma_1^2 \exp\!\!\left(\!-\frac{1}{2} \frac{(t-t')^2}{\ell_1^2}\right), 
\end{equation*}
%
which means the function values $\bm{f}_1=\{f_1(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $\bm{K}_1$, where $\bm{K}_{1_{t,s}}=k_1(t,s)$, with $t,s=1,\dots,T$.

The component $f_2(t)$ represents the yearly smooth seasonal pattern, using a periodic squared exponential covariance function (with period 365.25 to match the average length of the year) in a GP model,
%
\begin{equation*}
f_2(t) \sim \GP(0,k_2), \hspace{5mm} k_2(t,t') = \sigma_2^2 \exp\!\!\left(\!-\frac{2\,\sin^{\!2}\!(\pi(t-t')/365.25}{\ell_2^2}\right).
\end{equation*}
% which means the function values $\bm{f}_2=\{f_2(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_2$, where $K_{2_{t,s}}=k_2(t,s)$, with $t,s=1,\dots,T$.

The component $f_3(t)$ represents the weekly smooth pattern using a periodic squared exponential covariance function (with period 7 of length of the week) in a GP model,
%
\begin{equation*}
f_3(t) \sim \GP(0,k_3), \hspace{5mm} k_3(t,t') = \sigma_3^2 \exp\!\!\left(\!-\frac{2\,\sin^{\!2}\!(\pi(t-t')/7}{\ell_3^2}\right). 
\end{equation*}
% which means the function values $\bm{f}_3=\{f_3(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_3$, where $K_{3_{t,s}}=k_3(t,s)$, with $t,s=1,\dots,T$.

The component $f_4(t)$ represents the special days effects, modeled as a horse-shoe prior model \citep{carvalho2010,piironen2017sparsity}:
%
\begin{equation*}
f_4(t) \sim \Normal(0,\lambda^2_t \tau^2), \hspace{5mm} \lambda^2_t \sim \mathcal{C}^{+}(0,1).
\end{equation*}
%
A horse-shoe prior allows for sparse distributed effects. Its global parameter $\tau$ pulls all the weights (effects) globally towards zero, while the thick half-Cauchy tails for the local scales $\lambda_t$ allow some of the weights to escape the shrinkage. Different levels of sparsity can be accommodated by changing the value of $\tau$: with large $\tau$ all the variables have very diffuse priors with very little shrinkage towards zero, but letting $\tau \rightarrow 0$ will shrink all the weights $f_4(t)$ to zero \citep{piironen2016hyperprior}. 
%More details about the specified priors for $\tau$ and $\lambda_t$ have been included in Appendix \ref{app_horse_shoe}.

@@

The year effects $f_2(t)$ and week effects $f_3(t)$ use a periodic covariance function  and thus do no fit under the main framework of the HSGP approximation covered in this paper. However, they do have a representation based on expanding periodic covariance functions into a series of stochastic resonators (Appendix~\ref{ch5_sec_periodic}). Thus, the functions $f_2(t)$ and $f_3(t)$ are approximated as in equation (\ref{ch5_eq_f_period}), with variance coefficients $\tilde{q}_j^2$ as in equation (\ref{ch5_eq_q_2}).
We use $J=10$ cosine terms. The lengthscales estimates $\hat{\ell}_2$ and $\hat{\ell}_3$, for the GP components $f_2(t)$ and $f_3(t)$, respectively, are bigger than the minimum lengthscale reported by Figure \ref{ch5_fig28_m_lscale_periodic} as function of the number of cosine terms $J$, which means that the approximations are good.

@@

The Stan model code for the approximate GP model of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Birthday-data}.

@@

As the WBC predictor values were strictly positive and highly skewed, logarithm transformation is used for it. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
